A properly labelled proof system for intuitionistic modal logics

### 2019-09-16 caleidoscope/caleidoscope.tex
bib-file + new section for admissibility of mon + completeness

\usepackage[margin=2cm]{geometry}
\usepackage[noxy]{virginialake}                        
\usepackage{lmodern}
\usepackage{amssymb,amsmath,amsthm} 
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{pgf}
\usepackage[all]{xy}
\usepackage{tikz}
%\usepackage{hyperref}
\usepackage{pgffor}

@@

Since their introduction in the 1980s by Gabbay~\cite{gabbay:96},
\emph{labelled proof calculi} have been widely used by proof theorist
to give sound, complete and cut-free deductive system to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hyper-sequents~\cite{avron:elc96} or nested
sequents~\cite{kashima,brunnler,pog} or 2-sequents~\cite{masini:apal92},
labelled calculi have the advantage of beeing more uniform and
accommodating more logics.

Standard labelled sequent calculi attach to every formula $A$ a label
$x$, witten as $\labels xA$, and additionally admit so called
\emph{relational atoms} of the form $\accs xy$ where $\rel$ is a
binary relation symbol. These calculi work best for logics with
standard Kripke semantics, as in this case the relation $\rel$ is used
to encode the accessibility relation in the Kripke models, and the
frame condition corresponding to the desired logic can be directly
encoded as inference rules. Prominent examples are classical modal
logics and intuitionistic propositional logic, where, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$), can be straightforwardly translated into the
inference rule
\begin{equation}
  \label{eq:Rtrans}
  \vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relation atoms, and $\Gamma$ and
$\Delta$ for sets of labelled formulas.

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in
ordinary Kripke frames for modal logics, and a relation $\le$ that
corresponds to the future relation in ordinary Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} corresponds to both
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  as they are equivalent in classical
  modal logic. However, in intuiotionistic modal logic they are not
  equivalent, and even though the logic $\ISfour$, i.e., the
  intuitionistic version of the modal logic $\Sfour$, contains both
  axioms, they can be added independently to the logic $\IK$ (the
  intuitionistic version of $\K$). However, the proof theory of these
  new logics has not been studied before, as no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \qquand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
  \end{equation}
  have already been studied
  in~\cite{Plotkin}.\todo{check: did they get that wrong in  \cite{Plotkin}?}
\item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer.}\todo{check terminology in Simpson} models, as
  already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as a very unsatisfactory situation\todo{check what he
    said exactly} (see also \cite{mar:str:indexed}).
\end{enumerate}
In order to address these two issues we propose here to extend the
notion of labelled sequent by also allowing relational atoms of the
form $\futs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
     \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

Besides that, there is another pleasant observation to make about our
system. Ordinary labelled sequent systems for intuitionistic modal
logic are single-conclusion~\cite{simpson:phd}. The same is true for
the corresponding nested sequent
systems~\cite{str:fossacs13,mar:str:aiml}. It is possible to express
Maehara style multiple-conclusion systems in nested
sequents~\cite{kuz:str:maehara}, and therefore also in ordinary
labelled sequents. However, also in these systems there are rules
($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
premise, even though this is not the case in labelled
systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:83} for
intuitionistic logic. Our system eliminates this
discrepancy. Consequently, every rule in our system is invertible, as
we never delete information.\lutz{check, is this really the case?}

@@

This paper is organized as follows, In the next section (Section~\ref{sec:intmod}) we recall the standard syntax and semantics of intuitionistic modal logics. Then, in Section~\ref{sec:system} we present our system for the intuitionistic modal logic $\IK$. In Sections~\ref{sec:soundness} and~\ref{sec:completeness}, we show soundness and completeness of the system with cut. The cut elimination theorem, proved in Section~\ref{sec:cutelim}, then entails soundness and completeness for the cut-free system. Finally, in Section~\ref{sec:ext} we discuss the possible extension to the system to capture other intuitionistic modal logics. \lutz{what about doing that directly in the system presentation, like Negri does it?}

\todo{maybe paragraph on related work? cite Maffezioli, Naibo and Negri~\cite{Maffezioli}.}

@@

We write $w \Vdash a$ if $a \in V(w)$, and by definition, we always have $w \Vdash \top$ and never that $w \Vdash \bot$. 
%
Then the relation is extended to all formulas by induction, following the rules for both intuitionistic and modal Kripke models:
\begin{equation}
\label{eq:kripke}
\parbox{30em}{%
	$w \Vdash A \AND B$ iff $w \Vdash A$ and $w \Vdash B$

	$w \Vdash A \OR B$ iff $w \Vdash A$ or $w \Vdash B$

	$w \Vdash A \IMP B$ iff for all $\lb{w'}$ with $\futs w{w'}$, if $w' \Vdash A$ then $w' \Vdash B$

	$w \Vdash \BOX A$ iff for all $\lb{w'}$ and $\lb u$ with $\futs w{w'}$ and $\accs {w'}u$, $u \Vdash A$ \hfill %$(\ast)$

	$w \Vdash \DIA A$ iff there exists a $u$ such that $\accs wu$ and $u \Vdash A$.

	%We write $w \not\Vdash A$  if it is not the case that $w\Vdash A$.\\
}
\end{equation}
\todo{fix typography}

We write $w \not\Vdash A$ if it is not the case that $w \Vdash A$, but contrarily to the classical case, we do not have $w \Vdash \neg A$ iff $w \Vdash A$ (since $\neg A$ is defined as $A \IMP \bot$).

From the monotonicity of the valuation function $V$, we get a monotonicity property for the relation:
\begin{proposition}(Monotonicity) 
	For any formula $A$ and for $\lb w, \lb{w'} \in W$, if $\futs w{w'}$ and $w \Vdash A$, then $w' \Vdash A$.
\end{proposition}

\begin{definition}
	A formula $A$ is \emph{satisfied} in a model $\M = \langle W, \rel, \le, V \rangle$, if for all $w \in W$ we have $w \Vdash A$.
	%
	A formula $A$ is \emph{valid} in a frame $\F = \langle W, \rel, \le \rangle$, if for all valuations $V$, $A$ is satisfied in $\langle W, R, \le, V \rangle$.
\end{definition}

@@

\begin{theorem}[Fischer-Servi~\cite{Fischer}, Plotkin and Stirling~\cite{Plotkin}]\label{thm:plotkin}
	A formula $A$ is a theorem of $\IK$ if and only if $A$ is valid in every bi-relational frame.
\end{theorem}

@@

In this section we present our properly labelled sequent proof system
for intuitionistic modal logics. The starting point is the notion of a
\emph{labelled formula} which is a pair $\labels xA$ of a label $\lb
x$ and a formula $\fm A$. The intended meaning is that $\fm A$ holds
at world $\lb x$. A \emph{relation atom} is either an expression
$\accs xy$ or $\futs xy$ where $\lb x$ and $\lb y$ are labels and the
intended meaning is that the corresponing relation holds between the
worlds $\lb x$ and $\lb y$. This is enough to define the notion of
\emph{(labelled) sequent}, which is a triple $\lseq\B\Left\Right$,
where $\B$ is a set of relation atoms, and $\Left$ and $\Right$ are
sets of labelled formulas, all written as list, separated by commas.

@@

Now we can present the inference rules for \emph{system $\labIKp$} for the logic $\IK$.
We obtained this system, shown in Figure~\ref{fig:labIKp}, as follows.
Our starting point was the multiple-conlusion nested sequent system \emph{\`a la} Maehara (as presented in~\cite{kuz:str:maehara}), which can straightforwardly translated into the labelled setting, and which yields the rules $\llabrn\bot$, $\llabrn\AND$, $\rlabrn\AND$, $\llabrn\OR$, $\rlabrn\OR$, $\llabrn\DIA$, and $\rlabrn\DIA$ as we show them in Figure~\ref{fig:labIKp}. However, this translation also yields the rules $\rn{id'}$, $\llabrn\IMP'$, and~$\llabrn\BOX'$:
\begin{equation}
  \label{eq:nomon}
  \vlinf{\rn{id}'}{}{\B, \Left, \labels{x}{a} \SEQ \Right, \labels{x}{a} }{}
  \quad
  \vliinf{\llabrn\IMP'}{}{\B,\Left, \labels{x}{A} \IMP B \SEQ \Right}{\B,\labels{x}{A \IMP B}, \Left \SEQ \Right, \labels{x}{A}}{\B, \Left, \labels{x}{B} \SEQ \Right}
  \quad
  \vlinf{\llabrn\BOX}{}{\B, xRz, \Left, \labels{x}{\BOX A} \SEQ \Right}{\B, xRz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right}
\end{equation}
that are not sufficient for a complete system, and we will see below
why. Before, let us first look at the rules $\rlabrn\IMP$ and
$\rlabrn\BOX$. In the multiple-conlusion nested sequent system
of~\cite{kuz:str:maehara}, these are the two rules that force
single-conclusion. This is the reason why in our system we need the
$\le$ in the relational atoms; we need to go to the future of the
current world.  In the Kripke-semantics in~\eqref{eq:kripke} the two
connectives $\IMP$ and $\BOX$ are the ones that need the
future-relation $\le$. This relation is reflexive and transitive. In
order to capture that in the proof system, we need to add the rules
$\rn{refl}$ and $\rn{trans}$.

Finally, in the semantics, the two relations $\rel$ and $\le$ are
strongly connected through the two conditions $\rn{F_1}$ and $\rn{F_2}$. These need to be reflected in the proof
system, which is done by the two rules $\rn{F_1}$ and $\rn{F_2}$.
%
However, these two rule create new labels, and in order to be
complete, the system needs the \emph{monotonicity} rule:
\begin{equation}
  \label{eq:mono}
  \vlinf{\rn{mon}}{}{\lseq{\B, x \le x'}{\Left, \labels{x}{A}}\Right}{
    \lseq{\B, x \le x'}{\Left, \labels{x}{A}, \labels{x'}{A}}\Right}
\end{equation}
Since this rule is a form of contraction, it would cause the same
problems as contraction in a cut elimination proof. Hence, it is
preferable to have a system in which this rule is admissible. This is
the reason why we have monotonicity incorporated in the rules
$\rn{id}$, $\llabrn\IMP$ and $\llabrn\BOX$ in Figure~\ref{fig:labIKp},
instead of using the rules in~\eqref{eq:nomon}.

@@

\begin{proof} [Proof of soundness]
	We need to prove that each sequent rule of our system $\labIKp$ is sound. We want to prove that for all model $\M$ that satisfies the premise, then $\M$ satifies the conclusion. As the proof is similar for each of the rules, we prove the soundness of some particular rules:

@@

		\item 
		If there exists a proof 
		$\vlderivation{\vlhtr{\DD}{\B, \Left, \labels{x}{\TOP} \SEQ \Right}}$ 
		then there exists a proof 
		$\vlderivation{\vlhtr{\DD^{\bm\TOP}}{\B, \Left \SEQ \Right}}$

>>>>

\usepackage[margin=2cm]{geometry}
\usepackage[noxy]{virginialake}                        
\usepackage{lmodern}
\usepackage{amssymb,amsmath,amsthm} 
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{pgf}
\usepackage{array}
\usepackage[all]{xy}
\usepackage{tikz}
%\usepackage{hyperref}
\usepackage{pgffor}

@@

Since their introduction in the 1980s by Gabbay~\cite{gabbay:96},
\emph{labelled proof calculi} have been widely used by proof theorist
to give sound, complete and cut-free deductive system to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hyper-sequents~\cite{avron:elc96} or nested
sequents~\cite{kashima:sl94,brunnler:aml09,poggiolesi:tmp09} or 2-sequents~\cite{masini:apal92},
labelled calculi have the advantage of beeing more uniform and
accommodating more logics.

Standard labelled sequent calculi attach to every formula $A$ a label
$x$, witten as $\labels xA$, and additionally admit so called
\emph{relational atoms} of the form $\accs xy$ where $\rel$ is a
binary relation symbol. These calculi work best for logics with
standard Kripke semantics, as in this case the relation $\rel$ is used
to encode the accessibility relation in the Kripke models, and the
frame condition corresponding to the desired logic can be directly
encoded as inference rules. Prominent examples are classical modal
logics and intuitionistic propositional logic, where, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$), can be straightforwardly translated into the
inference rule
\begin{equation}
  \label{eq:Rtrans}
  \vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relation atoms, and $\Gamma$ and
$\Delta$ for multi-sets of labelled formulas.

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in
ordinary Kripke frames for modal logics, and a relation $\le$ that
corresponds to the future relation in ordinary Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} corresponds to both
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  as they are equivalent in classical
  modal logic. However, in intuiotionistic modal logic they are not
  equivalent, and even though the logic $\ISfour$, i.e., the
  intuitionistic version of the modal logic $\Sfour$, contains both
  axioms, they can be added independently to the logic $\IK$ (the
  intuitionistic version of $\K$). However, the proof theory of these
  new logics has not been studied before, as no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \qquand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
  \end{equation}
  have already been studied
  in~\cite{plotkin:stirling:86}.\todo{check: did they get that wrong in  \cite{plotkin:stirling:86}?}
\item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer.}\todo{check terminology in Simpson} models, as
  already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as a very unsatisfactory situation\todo{check what he
    said exactly} (see also \cite{mar:str:tableaux17}).
\end{enumerate}
In order to address these two issues we propose here to extend the
notion of labelled sequent by also allowing relational atoms of the
form $\futs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
     \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

Besides that, there is another pleasant observation to make about our
system. Ordinary labelled sequent systems for intuitionistic modal
logic are single-conclusion~\cite{simpson:phd}. The same is true for
the corresponding nested sequent
systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
Maehara style multiple-conclusion systems in nested
sequents~\cite{kuz:str:maehara}, and therefore also in ordinary
labelled sequents. However, also in these systems there are rules
($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
premise, even though this is not the case in labelled
systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:83} for
intuitionistic logic. Our system eliminates this
discrepancy. Consequently, every rule in our system is invertible, as
we never delete information.\lutz{check, is this really the case?}

@@

This paper is organized as follows, In the next section (Section~\ref{sec:intmod}) we recall the standard syntax and semantics of intuitionistic modal logics. Then, in Section~\ref{sec:system} we present our system for the intuitionistic modal logic $\IK$. In Sections~\ref{sec:soundness} and~\ref{sec:completeness}, we show soundness and completeness of the system with cut. The cut elimination theorem, proved in Section~\ref{sec:cut-elim}, then entails soundness and completeness for the cut-free system. Finally, in Section~\ref{sec:ext} we discuss the possible extension to the system to capture other intuitionistic modal logics. \lutz{what about doing that directly in the system presentation, like Negri does it?}

\todo{maybe paragraph on related work? cite Maffezioli, Naibo and Negri~\cite{maffezioli:etal:synthese13}.}

@@

We write $\M, w \Vdash a$ if $a \in V(w)$, and by definition, we always have $\M, w \Vdash \top$ and never that $\M, w \Vdash \bot$. 
%
Then the relation is extended to all formulas by induction, following the rules for both intuitionistic and modal Kripke models:
$$
\begin{array}{rcl}
\label{eq:kripke}
%\parbox{30em}{%

	\M, \lb w \Vdash \fm{A \AND B} & \mbox{iff} & \M, \lb w \Vdash \fm A \mbox{ and } \M, \lb w \Vdash \fm B\\

	\M, \lb w \Vdash \fm{A \OR B} & \mbox{iff} & \M, \lb w \Vdash \fm A \mbox{ or } \M, \lb w \Vdash \fm B\\

	\M, \lb w \Vdash \fm{A \IMP B} & \mbox{iff} & \mbox{for all } \lb{w'} \mbox{ with } \futs w{w'}, \mbox{ if } \M, \lb{w'} \Vdash \fm A \mbox{ then } \M, \lb{w'} \Vdash \fm B\\

	\M, \lb w \Vdash \fm{\BOX A} & \mbox{iff} & \mbox{for all } \lb{w'} \mbox{ and } \lb u \mbox{ with } \futs w{w'} \mbox{ and } \accs {w'}u, \M, \lb u \Vdash \fm A \hfill \\ %$(\ast)$

	\M, \lb w \Vdash \fm{\DIA A} & \mbox{iff} & \mbox{there exists a } \lb u \mbox{ such that } \accs wu \mbox{ and } \M, \lb u \Vdash \fm A.

	%We write $w \not\Vdash A$  if it is not the case that $w\Vdash A$.\\
%}
\end{array}
$$

We write $\M, \lb w \not\Vdash \fm A$ if it is not the case that $\M, \lb w \Vdash \fm A$, but contrarily to the classical case, we do not have $\M, \lb w \Vdash \fm{\neg A}$ iff $\M, \lb w \Vdash \fm A$ (since $\fm{\neg A}$ is defined as $\fm{A \IMP \bot}$).

From the monotonicity of the valuation function $V$, we get a monotonicity property for the relation:
\begin{proposition}(Monotonicity) 
	For any formula $\fm A$ and for $\lb w, \lb{w'} \in W$, if $\futs w{w'}$ and $\M, \lb w \Vdash \fm A$, then $\M, \lb{w'} \Vdash \fm A$.
\end{proposition}

\begin{definition}
	A formula $\fm A$ is \emph{satisfied} in a model $\M = \langle W, \rel, \le, V \rangle$, if for all $\lb w \in W$ we have $\M, \lb w \Vdash \fm A$.
	%
	A formula $\fm A$ is \emph{valid} in a frame $\F = \langle W, \rel, \le \rangle$, if for all valuations $V$, $\fm A$ is satisfied in $\langle W, R, \le, V \rangle$.
\end{definition}

@@

\begin{theorem}[Fischer-Servi~\cite{fischer-servi:84}, Plotkin and Stirling~\cite{plotkin:stirling:86}]\label{thm:plotkin}
	A formula $\fm A$ is a theorem of $\IK$ if and only if $\fm A$ is valid in every bi-relational frame.
\end{theorem}

@@

In this section we present our properly labelled sequent proof system
for intuitionistic modal logics. The starting point is the notion of a
\emph{labelled formula} which is a pair $\labels xA$ of a label $\lb
x$ and a formula $\fm A$. The intended meaning is that $\fm A$ holds
at world $\lb x$. A \emph{relation atom} is either an expression
$\accs xy$ or $\futs xy$ where $\lb x$ and $\lb y$ are labels and the
intended meaning is that the corresponing relation holds between the
worlds $\lb x$ and $\lb y$. This is enough to define the notion of
\emph{(labelled) sequent}, which is a triple $\lseq\B\Left\Right$,
where $\B$ is a set of relation atoms, and $\Left$ and $\Right$ are
multi-sets of labelled formulas, all written as list, separated by commas.

@@

Now we can present the inference rules for \emph{system $\labIKp$} for the logic $\IK$.
We obtained this system, shown in Figure~\ref{fig:labIKp}, as follows.
Our starting point was the multiple-conlusion nested sequent system \emph{\`a la} Maehara (as presented in~\cite{str:2017maehara}), which can straightforwardly translated into the labelled setting, and which yields the rules $\llabrn\bot$, $\llabrn\AND$, $\rlabrn\AND$, $\llabrn\OR$, $\rlabrn\OR$, $\llabrn\DIA$, and $\rlabrn\DIA$ as we show them in Figure~\ref{fig:labIKp}. However, this translation also yields the rules $\rn{id'}$, $\llabrn\IMP'$, and~$\llabrn\BOX'$:
\begin{equation}
  \label{eq:nomon}
  \vlinf{\rn{id}'}{}{\B, \Left, \labels{x}{a} \SEQ \Right, \labels{x}{a} }{}
  \quad
  \vliinf{\llabrn\IMP'}{}{\B,\Left, \labels{x}{A} \IMP B \SEQ \Right}{\B,\labels{x}{A \IMP B}, \Left \SEQ \Right, \labels{x}{A}}{\B, \Left, \labels{x}{B} \SEQ \Right}
  \quad
  \vlinf{\llabrn\BOX}{}{\B, xRz, \Left, \labels{x}{\BOX A} \SEQ \Right}{\B, xRz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right}
\end{equation}
that are not sufficient for a complete system, and we will see below
why. Before, let us first look at the rules $\rlabrn\IMP$ and
$\rlabrn\BOX$. In the multiple-conlusion nested sequent system
of~\cite{str:2017maehara}, these are the two rules that force
single-conclusion. This is the reason why in our system we need the
$\le$ in the relational atoms; we need to go to the future of the
current world.  In the Kripke-semantics in~\eqref{eq:kripke} the two
connectives $\IMP$ and $\BOX$ are the ones that need the
future-relation $\le$. This relation is reflexive and transitive. In
order to capture that in the proof system, we need to add the rules
$\rn{refl}$ and $\rn{trans}$.

Finally, in the semantics, the two relations $\rel$ and $\le$ are
strongly connected through the two conditions $\rn{F_1}$ and $\rn{F_2}$. These need to be reflected in the proof
system, which is done by the two rules $\rn{F_1}$ and $\rn{F_2}$.
%
However, these two rule create new labels, and in order to be
complete, the system needs the \emph{monotonicity} rule:
\begin{equation}
  \label{eq:mono}
  \vlinf{\rn{mon}}{}{\lseq{\B, x \le x'}{\Left, \labels{x}{A}}\Right}{
    \lseq{\B, x \le x'}{\Left, \labels{x}{A}, \labels{x'}{A}}\Right}
\end{equation}
Since this rule is a form of contraction, it would cause the same
problems as contraction in a cut elimination proof. Hence, it is
preferable to have a system in which this rule is admissible. This is
the reason why we have monotonicity incorporated in the rules
$\rn{id}$, $\llabrn\IMP$ and $\llabrn\BOX$ in Figure~\ref{fig:labIKp},
instead of using the rules in~\eqref{eq:nomon}. We prove the admissibility of monotonicity in the Section~\ref{sec:mon}.

\todo{I am not sure how to introduce the next lemma}

\begin{proof}
	By induction of the size of the formula $\fm A$.
	\begin{itemize}
		\item Atomic case
		\begin{smallequation*}
			%\vlinf{\labrn{id_g}}{}{\B, x \le y; \Left, \labels{x}{a} \SEQ \Right, \labels{y}{a}}{}
			%\reducesto
			\vlinf{\labrn{id}}{}{\B, x \le y; \Left, \labels{x}{a} \SEQ \Right, \labels{y}{a}}{}
		\end{smallequation*}

		\item $\fm{A \AND B}$
		%\begin{smallequation*}
		%	\vlinf{\labrn{id_g}}{}{\B, x \le y; \Left, \labels{x}{A \AND B} \SEQ \Right, \labels{y}{A \AND B}}{}
		%	\reducesto
		%\end{smallequation*}
		\begin{smallequation*}
			\vlderivation{
				\vlin{\llabrn\AND}{}{\B, x \le y; \Left, \labels{x}{A \AND B} \SEQ \Right, \labels{y}{A \AND B}}{
					\vliin{\rlabrn\AND}{}{\B, x \le y; \Left, \labels{x}{A}, \labels{x}{B} \SEQ \Right, \labels{y}{A \AND B}}{
						\vlin{\labrn{id_g}}{}{\B, x \le y; \Left, \labels{x}{A}, \labels{x}{B} \SEQ \Right, \labels{y}{A}}{
							\vlhy{}
						}
					}{
					\vlin{\labrn{id_g}}{}{\B, x \le y; \Left, \labels{x}{A}, \labels{x}{B} \SEQ \Right, \labels{y}{B}}{
						\vlhy{}
					}
				}
			}
		}
	\end{smallequation*}

	\item $\fm{A \OR B}$
	%\begin{smallequation*}
	%	\vlinf{\labrn{id_g}}{}{\B, x \le y; \Left, \labels{x}{A \OR B} \SEQ \Right, \labels{y}{A \OR B}}{}
	%	\reducesto
	%\end{smallequation*}
	\begin{smallequation*}
		\vlderivation{
			\vliin{\llabrn\OR}{}{\B, x \le y; \Left, \labels{x}{A \OR B} \SEQ \Right, \labels{y}{A \OR B}}{
				\vlin{\rlabrn\OR}{}{\B, x \le y; \Left, \labels{x}{A} \SEQ \Right, \labels{y}{A \OR B}}{
					\vlin{\labrn{id_g}}{}{\B, x \le y; \Left, \labels{x}{A} \SEQ \Right, \labels{y}{A}}{
						\vlhy{}
					}
				}
			}{
			\vlin{\rlabrn\OR}{}{\B, x \le y; \Left, \labels{x}{B} \SEQ \Right, \labels{y}{A \OR B}}{
				\vlin{\labrn{id_g}}{}{\B, x \le y; \Left, \labels{x}{B} \SEQ \Right, \labels{y}{B}}{
					\vlhy{}
				}
			}
		}
	}
	\end{smallequation*}

	\item $\fm{A \IMP B}$
%	\begin{smallequation*}
%		\vlinf{\labrn{id_g}}{}{\B, x \le y; \Left, \labels{x}{A \IMP B} \SEQ \Right, \labels{y}{A \IMP B}}{}
%		\reducesto
%	\end{smallequation*}
	\begin{smallequation*}
		\vlderivation{
			\vlin{\rlabrn\IMP}{}{\B, x \le y; \Left, \labels{x}{A \IMP B} \SEQ \Right, \labels{y}{A \IMP B}}{
				\vlin{\rn{trans}}{}{\B, x \le y, y \le z; \Left, \labels{x}{A \IMP B}, \labels{z}{A} \SEQ \Right, \labels{z}{B}}{
					\vliin{\llabrn\IMP}{}{\B, x \le y, y \le z, x \le z; \Left, \labels{x}{A \IMP B}, \labels{z}{A} \SEQ \Right, \labels{z}{B}}{
						\vlin{\rn{refl}}{}{\B, x \le y, y \le z, x \le z; \Left, \labels{x}{A \IMP B}, \labels{z}{A} \SEQ \Right, \labels{z}{B}, \labels{z}{A}}{
							\vlin{\labrn{id_g}}{}{\B, x \le y, y \le z, x \le z, z \le z; \Left, \labels{x}{A \IMP B}, \labels{z}{A} \SEQ \Right, \labels{z}{B}, \labels{z}{A}}{
								\vlhy{}
							}
						}
					}{
					\vlin{\rn{refl}}{}{\B, x \le y, y \le z, x \le z; \Left, \labels{z}{B}, \labels{z}{A} \SEQ \Right, \labels{z}{B}}{
						\vlin{\labrn{id_g}}{}{\B, x \le y, y \le z, x \le z, z \le z; \Left, \labels{z}{B}, \labels{z}{A} \SEQ \Right, \labels{z}{B}}{
							\vlhy{}
						}
					}
				}
			}
		}
	}
	\end{smallequation*}

	\item $\fm{\BOX A}$
%	\begin{smallequation*}
%		\vlinf{\labrn{id_g}}{}{\B, x \le y; \Left, \labels{x}{\BOX A} \SEQ \Right, \labels{y}{\BOX A}}{}
%		\reducesto
%	\end{smallequation*}
	\begin{smallequation*}
		\vlderivation{
			\vlin{\rlabrn\BOX}{}{\B, x \le y; \Left, \labels{x}{\BOX A} \SEQ \Right, \labels{y}{\BOX A}}{
				\vlin{\rn{trans}}{}{\B, x \le y, y \le z, z \rel w; \Left, \labels{x}{\BOX A} \SEQ \Right, \labels{w}{A}}{
					\vlin{\llabrn\BOX}{}{\B, x \le y, y \le z, x \le z, z \rel w; \Left, \labels{x}{\BOX A} \SEQ \Right, \labels{w}{A}}{
						\vlin{\rn{refl}}{}{\B, x \le y, y \le z, x \le z, z \rel w; \Left, \labels{z}{\BOX A}, \labels{w}{A} \SEQ \Right, \labels{w}{A}}{
							\vlin{\labrn{id_g}}{}{\B, x \le y, y \le z, x \le z, z \rel w, w \le w; \Left, \labels{z}{\BOX A}, \labels{w}{A} \SEQ \Right, \labels{w}{A}}{
								\vlhy{}
							}
						}
					}
				}
			}
		}
	\end{smallequation*}

	\item $\fm{\DIA A}$
%	\begin{smallequation*}
%		\vlinf{\labrn{id_g}}{}{\B, x \le y; \Left, \labels{x}{\DIA A} \SEQ \Right, \labels{y}{\DIA A}}{}
%		\reducesto
%	\end{smallequation*}
	\begin{smallequation*}
		\vlderivation{
			\vlin{\llabrn\DIA}{}{\B, x \le y; \Left, \labels{x}{\DIA A} \SEQ \Right, \labels{y}{\DIA A}}{
				\vlin{\color{red}{\rn{F_2}}}{}{\B, x \le y, x \rel z; \Left, \labels{z}{A} \SEQ \Right, \labels{y}{\DIA A}}{
					\vlin{\rlabrn\DIA}{}{\B, x \le y, x \rel z, z \le u, y \rel u; \Left, \labels{z}{A} \SEQ \Right, \labels{y}{\DIA A}}{
						\vlin{\labrn{id_g}}{}{\B, x \le y, x \rel z, z \le u, y \rel u; \Left, \labels{z}{A} \SEQ \Right, \labels{y}{\DIA A}, \labels{u}{A}}{
							\vlhy{}
						}
					}
				}
			}
		}
	\end{smallequation*}
	\end{itemize}

\section{Admissibility of monotonicity}\label{sec:mon}

@@

\begin{proof} [Proof of soundness]
	We need to prove that each sequent rule of our system $\labIKp$ is sound. We want to prove that for all model $\M$ that satisfies the premise, then $\M$ satifies the conclusion.

	In order to obtain the proof, let us introduce a definition:

	\begin{definition}

	\end{definition}

	As the proof is similar for each of the rules, we prove the soundness of some particular rules:

@@

	%	\item 
	%	If there exists a proof 
	%	$\vlderivation{\vlhtr{\DD}{\B, \Left, \labels{x}{\TOP} \SEQ \Right}}$ 
	%	then there exists a proof 
	%	$\vlderivation{\vlhtr{\DD^{\bm\TOP}}{\B, \Left \SEQ \Right}}$
